---
title: "Open water flow"
output: 
  html_notebook:
   css:  want_n.css
author: G. Bier
---

# Introduction

This document is designed to give a short introduction in some of the hydraulic properties of channel flow and some basic concept to calculate water levels in rivers, canals, channels etc..
It is NOT intended to be a crash course in hydraulics.

# Equations for open water flow

The flow in rivers as for example the Regge is most often described by the so called St-Venant equations, expressing respectively the mass and the momentum balance (only in *x*-direction):

$$
\begin{align*}
\frac{\partial A}{\partial t} + \frac{\partial Q}{\partial x} &= I &\text{: mass balance}\\
\frac{\partial Q}{\partial t} + \frac{\partial Q\,u}{\partial x} 
            &= g\;A\;\Big( S_o - S_f - \frac{\partial a}{\partial x}\Big) &\text{: momentum balance}
\end{align*}
$$

As we are here less interested in highly dynamic open water calculations, a simplification of the second equation above will be used. This simplification starts from the observation that in many situations (certainly lowland situations) the first two terms in the momentum part of the St-Venant equations are much smaller than the terms on the right hand side.

$$
\frac{\partial Q}{\partial t} + \frac{\partial Q\,u}{\partial x} = 
        g\;A\;\Big( S_o - S_f - \frac{\partial a}{\partial x}\Big)\\
\Downarrow \\
0 = g\;A\;\Big( S_o - S_f - \frac{\partial a}{\partial x}\Big)\\
\Downarrow \\
0 =  S_o - S_f - \frac{\partial a}{\partial x}
$$

gb

Volgens Paul "openwater.pdf" van 2007.

Stationary formulas, including the side inflow

$$
 \frac{\partial Q}{\partial s}=I \\
 \frac{\partial Qu}{\partial s}=g \, A \, \{ S_o+S_f+S_p \}
$$

Rewriting the second equation

$$
u = Q/A ,\,\; \partial A = b\, \partial a ??\\
\frac{\partial Qu}{\partial s}= \frac{\partial}{\partial s}\left ( \frac{Q^2}{A}\right) = \\
2\frac{Q}{A}\frac{\partial Q}{\partial s}-\frac{Q^2}{A^2}\frac{\partial A}{\partial s} = \\
2uI - \frac{Q^2b}{A^2}\frac{\partial a}{\partial s}
$$
Putting the last derived (rewritten) equation into the previous formula block:

$$
\frac{\partial Qu}{\partial s}=g \, A \, \{ S_o+S_f+S_p \} \\
2uI - \frac{Q^2b}{A^2}\frac{\partial a}{\partial s} = g \, A \, \{ S_o+S_f+ \frac{\partial a}{\partial s} \} \\
\left ( 1- \frac{Q^2b}{gA^3}\right ) \frac{\partial a}{\partial s} = S_o - S_f - \frac{2uI}{gA} \\
\Leftrightarrow \frac{\partial a}{\partial s} = \frac{S_o - S_f -S_i}{1-Fr^2}
$$


The meaning of the terms in these equations will be explained when they first appear in this document.

In these exercises situations will be studied where

-   the flow in the Hooge Raam can be considered to be *stationary* (so the time derivatives in the equations above disappear)
-   there is a weir at the downstream end of the river

Situations of this type are often called **backwater curves**.

This type of open water model that will be developed in this document will be coupled with a groundwater model. So at some places typical groundwater terms and dimensions may occur.

The open water has its own standard units:

-   length unit = $\bf{m}$ meter
-   time unit = $\bf{s}$ second

## Geometry of the open water domain

In contrast to groundwater flow, open water flow is limmitid to the open water course itself. This could be simulated in 2D, in de cross sectional and depth domain, or more frequently use averged depth. As can be seen in the abovementionded equations.

The cross sectional geometry of the open water course is of major importance since the hydraulic properties of A, P and R change w.r.t. the water level. Therefor extra intention is paid on this aspect. 
Nowaday many "parenthesis" shaped profiles are used to accommodate base flow and highly discharge events without drastically changing open water levels.
Here we will consider, for simplicity, only trapezoid profiles.
Although simplified, these profiles have non-linear behaviour regarding A,P and R.

Below an illustration of such a trapezoid profile:


$A$ is the wetted Area of the profile, $P$ being the wettted perimeter (the lenght of the profile which is wet due to the water level) and $R$, being the hydraulic radius and determined by: $R=\frac{A}{P}$.

<div class="question">
Suppose this profile has a side slope `m` of 1.5, what are the A,P and R relations as a function of the water level.
</div>

## Wetted area, perimeter and radius of the open water course

One of the important cross section functions is the wetted area. In the St-Venant equation denoted by $A$. For a trapezoidal cross section this can be easily calculated by the following function:

$$
A = b\,a + m\,a^2 \\
P = b + 2\sqrt{1+m^2}\,a \\
R = \frac{A}{P} \\
B = b + 2m\,a
$$

```{r, eval=TRUE}

rm(list = ls())


ow.b = 1 #just a guess
ow.m = 1 #just a guess

##Upstream Discharge 
#Qup = 1.25 #m3/s
Qup = 1.25 + 1.60125 #m3/s with side inflow based on the catchment area of the Regge and a normative discharge of 7 mm/d
ow.A = function(a)
{
  return(ow.b*a+ow.m*2*a^2/2)
}

a = seq(0,5, by =0.1)

plot(a,ow.A(a),main = "Wetted area")
grid()


ow.P = function(a)
{
  return(ow.b + a * 2 *sqrt(1 + ow.m^2))
}

plot(a,ow.P(a),main = "Wetted perimeter")
grid()

plot(a,ow.A(a)/ow.P(a),main = "Hydraulic radius")
grid()

ow.R = function(a)
{
  return(ow.A(a)/ow.P(a))
}

ow.B = function(a)
{
  return(ow.b + 2*ow.m*a)
}

```

## Bed slope

The open water course will have a slope $S_0$ starting upstream with a higher elevation than downstream. From a perspective of the river this is called a longitudinal profile.  
The slope is calculated as:

$$
S_0 = \frac{dZ}{ds}
$$
With $Z$ being the bottom elevation relative to a datum, usually NAP in the Netherlands. The $s$ stands for the length along the river.



Weir

$$
 Q = \frac{2}{3}\sqrt{2g} C_D b (h - d)^{3/2}
$$
With $C_D$ being the discharge coefficient, for a simple rectangular weir; 0.62
 

```{r}
S0 = 0.001 # about 0.5 m/km for the midden Regge


##total length of the river
ow.L = 5800 # distance (about between convulation point Exsooshe Aa and boven Regge till weir at Notter)
Zb.fun = approxfun(c(0,-ow.L),c(0,ow.L*S0))
```



## Equilibrium flow

$$
Q_\text{equi} = \frac{1}{n}\sqrt{S_0}AR^{2/3}
$$


```{r}
n = 0.05
a.range = seq(0.1,2,by = 0.1)
Qequi.fun = function(a)
{
  return(1/n * sqrt(S0)* ow.A(a) * ow.R(a)^(2/3))
}

Qequi.range = Qequi.fun(a.range)
plot(Qequi.range,a.range,type = "o", xlab = "Qequi (m3/s)",ylab = "a water depth (m)",
     main = "Equilibrium discharge")
grid()


##finding the equilibrium water depth for a given discharge
a.equilibrium = uniroot(function(a) Qequi.fun(a)-Qup,c(0.1,2))$root
print(paste("the equilibrium water level is : ",a.equilibrium, "for a discharge of ",Qup))

```

## Calculating $\frac{da}{ds}$ without side inflow



$$
 \frac{\partial a}{\partial s} = \frac{S_0 - S_f}{1-Fr^2} \\
 S_f = \frac{n^2 |Q|^2}{A^2R^{4/3}} \\
 Fr = \sqrt{\frac{Q^2B}{gA^3}}
$$

### Numerical approach

The calculation of the water level in the river can be done by a explicit scheme. 
So:

$$
\frac{\partial a}{\partial s} = \frac{S_0 - S_f}{1-Fr^2} \\
\Downarrow \\
\frac{\Delta a}{\Delta s} \approx \frac{S_0 - S_f}{1-Fr^2} \\
\Downarrow \\
a[s+\Delta s] = a[s ] + \frac{S_0 - S_f}{1-Fr^2} \Delta s
$$
Where $S_f$ and $Fr$ are calculated with the water depth at the current location $a[s]$.

### The Code 

To determine the $\frac{\Delta a}{\Delta s}$, code starts at the downstream end, where (for now) the water level is prescribed. The water depth $\Delta s$ upstream, $a[s + \Delta s]$ is based on $\,S_f(a[s]$ and $\,Fr^2(a[s])$. 
This can be accomplished using a loop, starting at the downstream end and moving upstream with a $\Delta s$ spatial step.
For all $a[s]$ till the before last location, which is $a[s=L]$ do:

  * Calculate $S_f$ and $Fr$ at the current location with the current $a[s]$
  * Calculate $\frac{\partial a}{\partial s}$ with $S_0$, $S_f$ and $Fr$
  * Calculate $a[s + \Delta s]$ with $a[s]$ and $\frac{\partial a}{\partial s}$  
  * Repeat the above steps till the before last location  

So for the last iteration through this loop, the before last gradient $\frac{\partial a}{\partial s}$ is used to calculate the water level at the last location, the upstream node.  


```{r}
#nodal distance
ds = -50
#trajectory of the river
s = seq(0,-ow.L, by = ds)
#create an array for the water depths
a.bw =s
# first downstream prescribed water level
a.bw[1] = 1.86
# Discharge rate upstream Qup
for (i in 1:(length(a.bw)-1) )
{
S.f = n^2*abs(Qup)^2/(ow.A(a.bw[i])^2*ow.R(a.bw[i])^(4/3))
Fr = sqrt(Qup^2*ow.B(a.bw[i])/(9.8*ow.A(a.bw[i])^3))
dads = (S0 - S.f)/(1-Fr^2)
##calculation of the water depths in upstream order
a.bw[i+1] = a.bw[i] + dads*ds
}
## the water level is simply the  water depts + the bottom slope
h.bw = a.bw + Zb.fun(s)

plot(s,h.bw, type = "l",ylim = c(0,max(h.bw)),xlab = "S (m)", ylab = "Water level (m)",
     main = paste("Backwater curve,S0",S0, "Qupstream",Qup),col = "blue",lwd = 3)
lines(s,Zb.fun(s), col = "brown", lwd = 3)
grid()
##plot the water level at equilibirum
lines(s,a.equilibrium + Zb.fun(s), col = "red", lty = 2, lwd = 3)
legend("bottomleft",legend = c("Backwater curve","Bottom elevation","Equilibrium water level"),
       col = c("blue","brown","red"),lty = c(1,1,2),lwd = c(3,3,3))
```

## Adding a weir

In de previous assignment the downstream water level was given. Now we will calculate the water level downstream of a weir. The weir is a simple rectangular weir with a discharge coefficient of 0.62. The weir is 1.5 m high and 2 m wide.

```{r}
#weir
Cd = 0.62

weir.d = 1.25 #weir/crest height
weir.b = 1.5 #

hweir = function(Q)
{
  return (weir.d + (Q/(2/3*sqrt(2*9.8)*Cd*weir.b))^(2/3))
}

## this function is more a control function to check the weir function
Qweir = function(h)
{
  ifelse(h < weir.d,0,2/3*sqrt(2*9.8)*Cd*weir.b*(h-weir.d)^(3/2))
 
}

####same chunck as above but now with a weir level based on Qup
# first downstream prescribed water level
a.bw[1] = hweir(Qup)
# Discharge rate upstream Qup
for (i in 1:(length(a.bw)-1) )
{
S.f = n^2*abs(Qup)^2/(ow.A(a.bw[i])^2*ow.R(a.bw[i])^(4/3))
Fr = sqrt(Qup^2*ow.B(a.bw[i])/(9.8*ow.A(a.bw[i])^3))
dads = (S0 - S.f)/(1-Fr^2)
##calculation of the water depths in upstream order
a.bw[i+1] = a.bw[i] + dads*ds
}
## the water level is simply the  water depts + the bottom slope
h.bw = a.bw + Zb.fun(s)

plot(s,h.bw, type = "l",ylim = c(0,max(h.bw)),xlab = "S (m)", ylab = "Water level (m)",
     main = paste("Backwater curve with weir,S0",S0, "Qupstream",Qup),col = "blue",lwd = 3)
lines(s,Zb.fun(s), col = "brown", lwd = 3)
grid()
##plot the water level at equilibirum
lines(s,a.equilibrium + Zb.fun(s), col = "red", lty = 2, lwd = 3)
legend("bottomleft",legend = c("Backwater curve","Bottom elevation","Equilibrium water level"),
       col = c("blue","brown","red"),lty = c(1,1,2),lwd = c(3,3,3))


```


## Froude is = 0 approach

For gradually flowing open water systems one could simplify to the Froude is = 0 approach.

$$
\frac{\partial a}{\partial s} = \frac{S_o - S_f}{1-Fr^2} \\
\Downarrow \\
Fr = \sqrt{\frac{Q^2B}{gA^3}} = 0 \\
\Downarrow \\
\frac{\partial a}{\partial s} = S_o - S_f \\
$$

```{r}
####same chunck as above but now with a weir level based on Qup
# first downstream prescribed water level
a.bw[1] = hweir(Qup)
a.bwf0 = a.bw
# Discharge rate upstream Qup
for (i in 1:(length(a.bw)-1) )
{
S.f = n^2*abs(Qup)^2/(ow.A(a.bw[i])^2*ow.R(a.bw[i])^(4/3))
Fr = sqrt(Qup^2*ow.B(a.bw[i])/(9.8*ow.A(a.bw[i])^3))
dads = (S0 - S.f)/(1-Fr^2)
dadsf0 = S0 - S.f
##calculation of the water depths in upstream order
a.bw[i+1] = a.bw[i] + dads*ds
a.bwf0[i+1] = a.bwf0[i] + dadsf0*ds
}
## the water level is simply the  water depth + the bottom slope
h.bw = a.bw + Zb.fun(s)
h.bwf0 = a.bwf0 + Zb.fun(s)

plot(s,h.bw, type = "l",ylim = c(0,max(h.bw)),xlab = "S (m)", ylab = "Water level (m)",
     main = paste("Backwater curve with weir,S0",S0, "Qupstream",Qup),col = "blue",lwd = 3)
lines(s,Zb.fun(s), col = "brown", lwd = 3)
grid()
##plot the water level at equilibirum
lines(s,a.equilibrium + Zb.fun(s), col = "red", lty = 2, lwd = 3)
lines(s,h.bwf0, col = "green", lty = 2, lwd = 3)
legend("bottomleft",legend = c("Backwater curve","Bottom elevation","Equilibrium water level","Froude is 0"),
       col = c("blue","brown","red","green"),lty = c(1,1,2),lwd = c(3,3,3))

##differences between fr0 and fr
plot(s,h.bw-h.bwf0, type = "l",xlab = "S (m)", ylab = "Difference in water level (m)",
     main = paste("Difference bw and bwfr=0,S0",S0, "Qupstream",Qup),col = "blue",lwd = 3)
grid()


```

## Adding side inflow S_i

In the previous derivation of some of the equations it is shown that 

$$
S_i = \frac{2uI}{gA}
$$

With:  

*   $S_i$ = side inflow $m/m$
*   $u$ = flow velocity $m/s$
*   $I$ = side inflow $m^2/s$

The local flow velocity (at $s[i]$) can be calculated as: $\frac{Q_s[i]}{A_s[i]}$ with $Q_s$ being the discharge at $s[i]$ and $A_s$ the wetted area at $s[i]$.

The side inflow can be calculated as a flux coming from local water courses like ditches. The normative discharge for grassland is 7-10 mm/d and for arable land 7-10 mm/d (ctt vademecum 1988). 

When the catchment area of the river is known, the side inflow can be calculated as: $I = \frac{Q_{catchment}}{L_{river}}$ with $Q$ being the total discharge entering the river from the catchment area over the length of the river $L_{river}$.

<div class="question">
Calculate the side inflow for the river Regge assuming a  catchment area of (5.8/29*9882ha; krw midden-regge) and a length of 5800 m.
</div>

<div class="answer">
Assume a normative discharge of 7 mm/d.  
```{r}
#The catchment area is:
A_catchm = 5.8/29*9882 *10000
# The discharge from side inflow:
Q_Si = A_catchm * 0.007/86400  # m3/s.
# The side inflow:
I = Q_Si/ow.L

```

</div>

To calculate the slope of the river with the side inflow, the following equation can be used:
$S_i = \frac{2uI}{gA}$. 
This means that besides $I$, the flow velocity $u$ and $A$ at each location along the river $s[i]$ is needed. The $u$ can be calculated as $u = \frac{Q}{A}$. The wetted area $A$ can be calculated as $A = b\,a + m\,a^2$, with $a$ at the current location.  
The local discharge is calculated starting at the weir and moving upstream. The local discharge is calculated as $Q_s[i] = Q_s[i-1] + I\,ds$. The local water depth is calculated as $a_s[i] = a_s[i-1] + \frac{S_0 - S_f - S_i}{1-Fr^2}\,ds$.



<div class="question">
Calculate the backwater curve now including the side inflow.
Simply copy the previous chunk and add the side inflow to the calculation of the water level.
</div>

<div class="answer"> 
```{r}
####same chunk as above but now side inflow Si is added
# Discharge at starting point of calculation, so at the weir
Qweir = Qup + I*ow.L
# setup the Qs and a.si array with zeros
Qs = rep(0,times = length(a.bw))
a.si = rep(0,times = length(a.bw))
# first downstream prescribed water level
a.si[1] = hweir(Qweir)
##COMPARING a.bw and a.si with both Qup = 1.25+1.60125
a.si[1] = hweir(Qup)
a.sif0 = a.si

#first Qs = Qweir
Qs[1] = Qweir
#loop for calculating the water levels
for (i in 2:(length(a.bw)-1) )
{
#calculate the Q, u and Si at the current location
Qs[i] = Qs[i-1] - I*-ds  ## ds is negative!! and I*ds should be subtracted for the Qs upstream
u = Qs[i]/ow.A(a.si[i-1])
S.i = 2*u*I/(9.8*ow.A(a.si[i-1]))
S.f = n^2*abs(Qs[i])^2/(ow.A(a.si[i-1])^2*ow.R(a.si[i-1])^(4/3))
Fr = sqrt(Qs[i]^2*ow.B(a.si[i-1])/(9.8*ow.A(a.si[i-1])^3))
dads = (S0 - S.f - S.i)/(1-Fr^2)
dadsf0 = S0 - S.f - S.i
##calculation of the water depths in upstream order
a.si[i] = a.si[i-1] + dads*ds
a.sif0[i] = a.sif0[i-1] + dadsf0*ds
}
## the water level is simply the  water depts + the bottom slope
h.si = a.si + Zb.fun(s)
h.sif0 = a.sif0 + Zb.fun(s)

plot(s,h.si, type = "l",ylim = c(0,max(h.si)),xlab = "S (m)", ylab = "Water level (m)",
     main = paste("Backwater curve with weir,S0",S0, "Qtotal",Qweir),col = "blue",lwd = 3)
lines(s,Zb.fun(s), col = "brown", lwd = 3)
grid()
##plot the water level at equilibirum
lines(s,a.equilibrium + Zb.fun(s), col = "red", lty = 2, lwd = 3)
lines(s,h.sif0, col = "green", lty = 2, lwd = 3)
legend("bottomleft",legend = c("Backwater curve","Bottom elevation","Equilibrium water level","Froude is 0"),
       col = c("blue","brown","red","green"),lty = c(1,1,2),lwd = c(3,3,3))

##differences between fr0 and fr
plot(s,h.si-h.sif0, type = "l",xlab = "S (m)", ylab = "Difference in water level (m)",
     main = paste("Difference bw.si and bwfr=0,S0",S0, "Qupstream",Qup),col = "blue",lwd = 3)
grid()

##plot h.bw without side inflow but with Qup=Qup+Qi and h.si with side inflow
plot(s,h.bw, type = "l",ylim = c(0,max(h.si)),xlab = "S (m)", ylab = "Water level (m)",
     main = paste("Backwater curve with and without side inflow,S0",S0, "Qupstream",Qup),col = "blue",lwd = 3)
lines(s,h.si, col = "red", lwd = 3)
grid()

##difference between the backwater curve with and without side inflow
plot(s,h.bw-h.si, type = "l",xlab = "S (m)", ylab = "Difference in water level (m)",
     main = paste("Difference bw and bw.si,S0",S0, "Qupstream",Qup),col = "blue",lwd = 3)
grid()

```

</div>