---
title: "Stationary Profiles Unsaturated Zone"
output: 
  html_notebook:
    css: want_n.css
---

# Background

## The Richards equation

The Richards equation is a partial differential equation that describes the flow of water in unsaturated soils. Flow in the unsaturated zone is predominantly vertical so here we only consider flow in the $z$ direction equation.

$$ \frac{\partial \theta}{\partial t} = \frac{\partial q}{\partial z} - S\\ \frac{\partial \theta}{\partial t} = \frac{\partial}{\partial z} \left\{ K(\psi) \left ( \frac{\partial \psi}{\partial z} +1 \right) \right\} -S $$

Where:

$\theta : \text{moisture content (L/L)}$

$q : \text{flux rate (L/T)}$

$S : \text{source and or sinks (L/T)}$

$K : \text{hydraulic conductivity (L/T)}$

$\psi : \text{pressure (L)}$

## The Retention curves

The moisture content and pressure head are closely related to each other. The retention curve is a plot of the moisture content as a function of the pressure head. The pressure head is negative in the unsaturated zone.

Some preliminaries to clean the environment and load soil data

```{r}
#cleasing the workspace
rm(list=ls())
#loading some packages
library(knitr)
#preprogrammed soil data
#loading the 
load("Staring.RData")
print(paste("Soil type: ",names(soil.set)," ",lapply(soil.set, function(x) x$name)))
#kable(paste("Soil type: ",names(soil.set)," ",lapply(soil.set, function(x) x$name)))
```

Plotting the retention curve for a specific soil type, here B13; Loam.

The first plot is based on the suction values (y-axis) vs. moisture content (x-axis).

The second, more usual plot, is based on the pF vs. moisture content. De pF stands for the log10 of the pressure values in centimeters.

So a pF of 2 means -100 cm pressure, pF=3 equals to -1000 cm pressure.

```{r}

#number of data points in pF range: 
pF = seq(0.01,4.2,by=0.05)
psi = -10^(pF)
soil = soil.set$B13
plot(soil$theta.fun(psi),psi,type="l",col="red",lwd = 3,
     main = paste("Retention profile of",soil$name),
     ylab="Pressure head (cm)",xlab="Water content at equilibrium")
grid()
plot(soil$theta.fun(psi),log10(-psi),lwd = 3, type="l",col="red",
     main = paste("Retention profile of",soil$name),
     ylab="Pressure head in pF",xlab="Water content at equilibrium")
grid()

# library(plotly)
# p <- plot_ly(x = ~soil$k.fun(psi), y = ~log10(-psi), type = 'scatter', mode = 'lines', 
#              line = list(color = 'red')) %>%
#   layout(title = 'Hydraulic Conductivity vs Pressure Head',
#          xaxis = list(title = 'Hydraulic conductivity'),
#          yaxis = list(title = 'Pressure head'))
# 
# # Toon de plot
# p


```

# The hydraulic conductivity

As shown in the Richards equation, the hydraulic conductivity is a function of pressure($K(\psi)$), which is negative in the unsaturated zone. However, it can also be expressed in terms of moisture content ($K(\theta)$) with the aid of the retention curve $\theta(\psi)$. [^1]

[^1]: Here we neglect effects of histeresis (soils turning from wet to dry do not describe the same moisture/pressure profiles as from soils turning dry from wet)

```{r}
## K as function of psi
plot(soil$k.fun(psi),psi,type="l",lwd = 3,col="red",main = paste("K(psi) of soil:",soil$name),
     ylab="Pressure head (cm)",xlab="Hydraulic conductivity")
grid()
plot(soil$k.fun(psi),log10(-psi),lwd = 3, type="l",col="red",
     main = paste("K(psi) of soil:",soil$name),
     ylab="Pressure head in pF",xlab="Hydraulic conductivity")
grid()

## K as function of theta 

plot(soil$theta.fun(psi),soil$k.fun(psi),type = "l",lwd = 3,col="orange", main = paste("K(theta) of soil:",soil$name),
     ylab ="Hydraulic conductivity (cm/d)",
     xlab = "Moisture content (m/m)")
grid()



```

### Comparing a few soils

```{r}
plot(soil.set$B13$theta.fun(psi),log10(-psi),type="l",col="red",lwd = 3,
     xlim = c(0.01,1),main = "Retention profiles",
     ylab="Pressure head (cm)",xlab="Water content at equilibrium")
lines(soil.set$B11$theta.fun(psi),log10(-psi),lwd = 3, col = "green")
lines(soil.set$O17$theta.fun(psi),log10(-psi), lwd = 3, col = "blue")
lines(soil.set$B5$theta.fun(psi),log10(-psi), lwd =3 , col = "brown")
grid()
##adding a legend to the plot
legend("topright", legend = c(soil.set$B13$name, soil.set$B11$name, soil.set$O17$name, soil.set$B5$name),
       col = c("red", "green", "blue", "brown"), lwd = 3)
```

```{r}
plot(soil.set$B13$k.fun(psi),log10(-psi),lwd = 3, type="l",col="red",
     main = paste("K(psi) of soil:",soil$name),
     xlim = c(0.0,soil.set$B5$ksat),
     ylab="Pressure head in pF",xlab="Hydraulic conductivity")
lines(soil.set$B11$k.fun(psi),log10(-psi),lwd = 3, col = "green")
lines(soil.set$O17$k.fun(psi),log10(-psi), lwd = 3, col = "blue")
lines(soil.set$B5$k.fun(psi),log10(-psi), lwd =3 , col = "brown")
grid()
##adding a legend to the plot
legend("topright", legend = c(soil.set$B13$name, soil.set$B11$name, soil.set$O17$name, soil.set$B5$name),
       col = c("red", "green", "blue", "brown"), lwd = 3)

```

# Stationary profiles

Here we will focus on the general behavior of flow in the unsaturated zone under stationary conditions for situations with deep and shallow groundwater tables and with different flux rates.

Since we only consider stationary flow, the time derivative is zero. The equation simplifies to:

$$
0 = \frac{\partial}{\partial z} \left\{ K(\psi) \left ( \frac{\partial \psi}{\partial z} +1 \right) \right\}
$$ And since this derivative $\frac{\partial}{\partial z}$ is zero it means that the outcome between the parenthesis, which is the flux rate $q$, is now a constant. The equation (Darcy Buckingham) simplifies to:

$$
q = K(\psi) \left ( \frac{\partial \psi}{\partial z} +1 \right)
$$

###  Procedure

1.  Define the soil type
2.  Define the domain of the model
3.  Define the boundary conditions
4.  Solve the equation

## Soil type

Choose one of the soil types given in the `soil.set` object.

## Domain

Since we are interested in the effect of large and small unsaturated zones the domains could be defined as one being 15 meters and the other 1 meter.

## Boundary conditions

On top of the domain, i.e. at the surface the flux rate will be given. A positive value indicates an influx, coming from precipitation for example. A negative value indicates an outflux, for example evaporation (bij soil and roots). At the bottom of the domain the pressure head is given. This is the pressure head of the groundwater table.

## Solve the equation

Here we are going to apply an explicit scheme to calculate the pressure head (i.e. $\psi$) in the unsaturated zone. The differential in the (Darcy like) equation is now approximated by a finite differences:

$$
\frac{\partial \psi}{\partial z} \approx \frac{\psi_{z}-\psi_{z-\Delta z}}{\Delta z}
$$

::: question
Rewrite the Darcy alike question into a finite difference equation that calculates the pressure head at depth $z$, based on all the other terms.
:::

::: answer
$$
\psi_{z} = \psi_{z-\Delta z} + \left( \frac{q_{z}}{K(\psi_{z-\Delta z})}  - 1 \right)\Delta z
$$
:::

## Implementation

Since the retetention curves are quite non-linear it is practical to have not too large steps ($\Delta z$) in the domain. Start with $\Delta z= 2 cm$

It is custom in the vadoze zone hydrology to use centimeters for the spatial scale. So the flux rate is given in cm/day. The hydraulic conductivity is given in cm/day. The pressure head is given in cm.

```{r}
#defining the soil type
#soil = soil.set$B13
soil = soil.set$O2
soil_name = soil$name

#defining the domain
dz = 1 #cm 
psi = seq(-1500,0,by=dz) #cm
z = psi

#boundary conditions

#at the groundwater table the pressure head psi is 0.0
psi[1] = 0.0

#at the surface the flux rate is given
q = 0.01 #cm/day



for (i in 2:length(psi)){
  
    psi[i] = psi[i-1] + (q/soil$k.fun(psi[i-1]) - 1)*dz
  
}

oldpar = par(no.readonly = TRUE)

par(mfrow=c(1,2)) #this creates two plots next to each other

plot(z,rev(z),type= "l",lty= "dashed",col="blue",main = paste("Soil:",soil_name,"; q = ",q,"cm/day"),
     ylab="Depth (cm)",xlab="Pressure head (psi in cm)",xlim=range(psi))
lines(psi,z,type="l",col="red")
grid()

range_theta = c(soil$theta.res,soil$theta.sat)
plot(soil$theta.fun(psi),z,type="l",col="red",
     ylab="Depth (cm)",xlab="Water content",xlim=range_theta)
grid()

par(oldpar)

```

Checking the flux rate

```{r}
#q = -k(psi)(dpsi/dz + 1))
q = soil$k.fun(psi)*(diff(psi)/dz + 1)
```
